Manuscript_draft

### 2021-02-27 manuscript_draft.Rmd
manuscript discussion

## Tree-level growth response to partial harvest

At the tree-level, our results suggest that while both hybrid white sruce and subalpine fir respond to thinning with faster radial growth, heavy thinning can elicit a stronger radial growth response in spruce, compared to fir.  This coincides with our finding of proportionally higher spruce volume and basal area in the heavy thinning treatment in 2019, compared to the other treatments.  Where thinning treatments in this region retain both species, removing a relatively high amount of the stand may favour spruce release.  Other studies have shown that residual spruce growth responds favourably to heavy thinning.  A 10-year analysis of white spruce growth underneath an aspen canopy found that a complete or high overstory removal resulted in immediate and significant spruce radial and height growth responses, compared to control and partial cutting treatments (Smith et al. 2016).   In a 35-year old white spruce stand, diameter increment was maximized with the heaviest thinning intensity, however, the treatment thinning from below (Stiell 1970).  Frank (1973) also showed that 70-year old white spruce basal increment increased with the amount of thinning. Competition for light is a key factor that limits white spruce growth, and opening up the stand to allow for increased light penetration may be a primary mechanism that enhances white spruce growth responses [@Lieffers1999].  High basal area removal may also increase air and soil temperatures, which can preferentially favour spruce growth responses compared to fir [@Lajzerowicz2004].

Other comparative studies of spruce and fir growth rates present differing evidence regarding species differences in release potential to thinning.  In mature Engelmann spruce (*Picea engelmannii*) – subalpine fir forests in western North America, understory spruce growth rates exceeded those of fir, allowing spruce to achieve canopy position faster, including after canopy gaps were created through disturbance (Antos et al. 2000, Andrus et al. 2018).  In mature forest in central BC, spruce growth exceeded that of subalpine fir after age 80 (Eis and Craigdallie 1983).  Because spruce tends to live longer than subalpine fir, higher growth rates at later ages may allow this species to become larger than fir.  In contrast, thinning studies in spruce-fir forests in eastern North America report that balsam fir (*Abies balsamifera*) responds to release with faster height and diameter growth than co-occurring red spruce (*Picea rubens*) (Dumais and Prevost, 2007; Dumais et al. 2014, Messier et al. 1999). Although both species are similar in shade tolerance, firs are considered to be better adapted to low light conditions in the understory, allowing them to persist longer in undisturbed forest  (Antos 2000) and potentially able to respond faster to changes in understory light (Dumais et al 2014, Messier et al. 1999).  Other studies have found no difference in growth response between spruce and fir species after treatment (MCaughey and Smith1982).   Our study also suggests that at lower levels of basal area removal, spruce and fir growth responses are similar.

In both species, post-treatment growth rates increased with tree size at the time of treatment.  This has been found in other studies as well (references) and may be explained by the phenomenon of size-asymmetric competition whereby larger trees are able to preempt resources, especially light, from smaller individuals [@Bose2017].  The implications of size-assymetric competition in thinned stands is that size inequality among trees may increase [@Cordonnier2015], and growth increases from treatment will be predominantly concentrated in the larger tree size classes [@Bose2018; @Schwinning1998].  Smaller trees will likely continue to experience competition and will not benefit as much from the thinning treatment.  Thinning treatments may not promote growth responses in the largest docminant trees in a stand either, as these trees may not experience significant changes in resource availability or may have achieved maximal growth rates [@Ryan1997, Bose2017]. Forest managers should account for potential size differential when planning thinning treatments.  For example, a silviculture treatment that seeks to release a suppressed understorey may have to ensure that sufficient overstorey trees are removed to reduce assymetric competition after treatment.

@@

Stand density 
The thinned treatments increased tree density faster than the control, and within three decades, there were no significant density differences between the three units.  Increasing stand density reflects the growth of smaller trees that over time, achieved the minimum diameter class (i.e., 7.5cm DBH) in this study.  The faster increases in stand density of the thinned treatments compared to the control likely reflect that smaller trees in the thinned stands responded with faster diameter increment after treatment (as previously discussed).  We note that our study did not account for new seedlings that may have become established over the period of the study.  Based on increasing densities in the smaller diameter classes, these stands will likely maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.  Thus, thinning treatments may be used in this context to maintain a structure typical of uneven-aged stands.

@@

Volume and basal area
Over the 27-year period of this study, the high-removal thinning treatments accrued basal area significantly faster than the control and low-removal thinning treatment, but not volume.  The volume formula used in this study uses both tree diameter and height [@Nigh2016].  Treatment-level differences in basal area increment but not volume suggest that tree growth responses to thinning might be primarily concentrated in radial growth, as opposed to height growth. It can be difficult to precisely measure tree height growth using field-based methods, particularly in mature trees, which may be a primary reason that studies tend to focus on radial growth responses to silviculture treatments and gap dynamics [@Vepakommaetal2011].    Enhanced radial growth is a common repsonse to reduced stand densities [@Plauborg2014].

Our results showed that the low-removal treatment unit and the control unit were similar in terms of stand volume and basal area, and that volume/basal area increment did not differ between the two.  The gaps created by the low-removal treatment may have been too small to elicit a strong growth response as observed in the high-removal thinning treatment.  Light levels may differ little between forest canopies that are closed or contain small gaps [@Bartemucci2002], and strong vertical light gradients may result in very little light regime changes for understorey trees.

@@

Over the 27-year period, the high-removal treatment unit accrued ca. 8m2 per hectare basal area more than the two other units.

Volume increment did not significantly differ between the units.

>>>>

## Tree- and stand-level growth response to partial harvest

### Species

At the tree-level, our results suggest that while both hybrid white spruce and subalpine fir respond to thinning with faster radial growth, heavy thinning can elicit a stronger radial growth response in spruce, compared to fir.  Spruce trees in the low-removal harvest treatment did not respond with a significant growth release.  THus, our results suggest that a relatively high removal of the stand will result in increased spruce growth, and this release will exceed that of co-occurring fir trees.  Other studies have shown that residual spruce growth responds favourably to heavy thinning.  A 10-year analysis of white spruce growth underneath an aspen canopy found that a complete or high overstory removal resulted in immediate and significant spruce radial and height growth responses, compared to control and partial cutting treatments (Smith et al. 2016).   In a 35-year old white spruce stand, diameter increment was maximized with the heaviest thinning intensity, however, the treatment thinning from below (Stiell 1970).  Frank (1973) also showed that 70-year old white spruce basal increment increased with the amount of thinning. Competition for light is a key factor that limits white spruce growth, and opening up the stand to allow for increased light penetration may be a primary mechanism that enhances white spruce growth responses [@Lieffers1999].  High basal area removal may also increase air and soil temperatures, which can preferentially favour spruce growth responses compared to fir [@Lajzerowicz2004].  The relatively low spruce growth response in the low-removal treatment may be explained in that the gaps created may have been too small to elicit as strong a growth response as observed in the high-removal thinning treatment.  Light levels may differ little between forest canopies that are closed or contain small gaps [@Bartemucci2002], and strong vertical light gradients may result in relatively little light regime changes for understorey trees.

In comparison, both partial harvest treatments stimulated a growth response in fir, compared to the control.  Although both species are similar in shade tolerance, firs are considered to be better adapted to low light conditions in the understory, allowing them to persist longer in undisturbed forest  (Antos 2000) and potentially able to respond faster to even modest changes in understory light (Dumais et al 2014, Messier et al. 1999).  Although fir growth may respond at lower levels of overstorey removal, the growth response potential of fir may be lower than spruce.  Comparative studies of spruce and fir growth rates present differing evidence regarding species differences in release potential to thinning.  In mature Engelmann spruce (*Picea engelmannii*) – subalpine fir forests in western North America, understory spruce growth rates exceeded those of fir, allowing spruce to achieve canopy position faster, including after canopy gaps were created through disturbance (Antos et al. 2000, Andrus et al. 2018).  In mature forest in central BC, spruce growth exceeded that of subalpine fir after age 80 (Eis and Craigdallie 1983).  Because spruce tends to live longer than subalpine fir, higher growth rates at later ages may allow this species to become larger than fir.  In contrast, thinning studies in spruce-fir forests in eastern North America report that balsam fir (*Abies balsamifera*) responds to release with faster height and diameter growth than co-occurring red spruce (*Picea rubens*) (Dumais and Prevost, 2007; Dumais et al. 2014, Messier et al. 1999). Other studies have found no difference in growth response between spruce and fir species after treatment (MCaughey and Smith1982).   Our study also suggests that at lower levels of basal area removal, spruce and fir growth responses are similar.

Post-treatment growth rates increased with tree size at the time of treatment, for both species.  This has been found in other studies as well (references) and may be related to size-asymmetric competition whereby larger trees preempt resources, especially light, from smaller individuals [@Bose2017].  Thinning treatments may exacerbate size-assymetric competition [@Cordonnier2015], and growth increases from treatment will be predominantly concentrated in the larger tree size classes [@Bose2018; @Schwinning1998].  Smaller trees will likely continue to experience competition and will not benefit as much from the thinning treatment.  Thinning treatments may not promote growth responses in the largest dominant trees in a stand either, as these trees may not experience significant changes in resource availability or may have achieved maximal growth rates [@Ryan1997, Bose2017]. Forest managers should account for potential size differential when planning thinning treatments.  For example, a silviculture treatment that seeks to release a suppressed understorey may have to ensure that sufficient overstorey trees are removed to reduce assymetric competition after treatment.

@@

Stand density 
The thinned treatments increased tree density faster than the control, and within three decades, there were no significant density differences between the two harvested treatments and control.  Increasing stand density reflects the growth of smaller trees that over time, achieved the minimum diameter class (i.e., 7.5cm DBH) in this study.  The faster increases in stand density of the thinned treatments compared to the control likely reflect that smaller trees in the thinned stands responded with faster diameter increment after treatment (as previously discussed).  We note that our study did not account for new seedlings that may have become established over the period of the study.  Based on increasing densities in the smaller diameter classes, these stands will likely maintain a negative-exponential diameter distribution, which is considered typical of older, uneven-aged stands in this region.  Thus, partial harvest treatments may be used in this context to maintain a structure typical of uneven-aged stands.

@@

### Stand-level growth responses
Over the 27-year period of this study, plots with high-removal harvest treatments accrued basal area significantly faster than the control and low-removal harvest treatment, likely due to increases in tree-level growth and stand density discussed previously.  As a result, differences in basal area between the high-removal harvest unit and the control reduced from ca. 17m2/hectare in 1992 to ca.9m2/hectare in 2019.  In contrast, basal area increment between the low-removal unit and the control did not differ significantly over the 1992-2019 period.  The difference between these two units remained relatively constant over all measurement periods at ca. 10m2/hectare.  The similar rates of basal area increment between the low-removal thinning treatment and the control are difficult to explain.  Tree-level results showed that fir grew faster in the low-removal unit than the control, but spruce did not. As well, by 2019 there were no significant differences in stand density between the two units.  Thus, it appears that while the fir trees may have increased growth as a response of low-removal, the growth increases were insufficient to cause a significant change in basal area increment.  Furhter monitoring in this study will help determine the future trajectory of the low-removal thinning treatment.

While basal area increment differed between the high-removal treatment unit and the control, there was little difference in volume increment between any of the units.  A possible explanation is that tree growth responses to thinning might be primarily concentrated in radial growth, as opposed to height growth; this would be important because the volume formula used in this study is based on both tree diameter and height [@Nigh2016].  It can be difficult to precisely measure tree height growth using field-based methods, particularly in mature trees, which may be a primary reason that studies tend to focus on radial growth responses to silviculture treatments and gap dynamics [@Vepakommaetal2011].  Commercial thinning treatments in spruce fir stands increased tree taper [@]    Enhanced radial growth is a common repsonse to reduced stand densities [@Plauborg2014].

@@
